---
title: "solution"
format: html
---

## Setup

### Attach required R packages

```{r}
#| message: false
library(batchelor)
library(edgeR)
library(scater)
library(ggplot2)
library(scran)
library(pheatmap)
library(scuttle)
```

### Load data

For convenience and reliability, the data were downloaded as described in the [Carpentries tutorial](https://carpentries-incubator.github.io/bioc-scrnaseq/multi-sample.html#setup-and-data-exploration) and saved locally as an RDS file.

```{r}
sce <- readRDS("/storage/olin0164/single_cell/pseudobulk/sce.rds")
```

## First steps

Preview the full `sce` object.

```{r}
sce
```

Preview the cell metadata.

```{r}
colData(sce)
```

## Pre-process the data

### Tidy up

Let's immediately switch the column 'sample' to a factor.
Aside from being good practice generally, it will also ensure that plots treat sample IDs as labels on a discrete scale, not values on a continuous scale

```{r}
sce$sample <- as.factor(sce$sample)
```

### Quality filtering

Remove stripped nuclei and doublets.

**Bonus point:**
To speed up downstream steps, randomly select 50% of the cells that remain after quality filtering in each sample.
In other words, if samples 5 and 6 had 100 and 200 cells after filtering, select 50 cells from sample 5 and 100 cells from sample 6.

```{r}
drop <- sce$celltype.mapped %in% c("stripped", "Doublet")

sce <- sce[,!drop]
```

### Normalisation

Normalise the data using `logNormCounts()`.
Re-assign the output to `sce`.

```{r}
sce <- logNormCounts(sce)
sce
```

If you preview the updated object, you should see a new assay named 'logcounts'.

### Model gene variance

Model gene variance (total, biological, technical) using `modelGeneVar()`.
Provide the sample of origin for each cell as the 'blocking' factor.
Assign the output to a new object called `dec`.

```{r}
dec <- modelGeneVar(sce, block = sce$sample)
dec
```

If you preview the object, you should see the typical statistics for variance modelling, plus a column called 'per.block' due to the use of the `block=` argument.
Refer to the help page of `modelGeneVar()` for detailed information.

### Select highly variable genes

Create a logical vector called `chosen.hvgs` that indicates whether the biological variance of each gene is strictly greater than 0.
Use `table()` to display the number of genes that meet that condition or not.

```{r}
chosen.hvgs <- dec$bio > 0
table(chosen.hvgs)
```

### Apply Principal Component Analysis

As a first dimensionality reduction step, use `runPCA()` to perform PCA with the following settings:

- Supply the set of highly variable features selected above, as the subset of rows to consider for PCA.
- Further refine that set by selecting only the 1,000 genes with the highest variances.

Re-assign the output to `sce`.

Read the help page of `scater::runPCA()` to identify the arguments implementing those settings.

```{r}
sce <- runPCA(sce, subset_row = chosen.hvgs, ntop = 1000)
sce
```

If you preview the object, you should see a new dimensionality reduction entry called 'PCA'.

### Apply t-distributed stochastic neighbor embedding

As a second dimensionality reduction step, use `runTSNE()` to perform Uniform t-distributed stochastic neighbor embedding on the PCA output.

Re-assign the output to `sce`.

```{r}
sce <- runTSNE(sce, dimred = "PCA")
sce
```

If you preview the object, you should see a new dimensionality reduction entry called 'TSNE'.

### Visualise the t-SNE layout

While we can use `plotTSNE()` for a quick preview, using `ggplot2` ourselves gives us more flexibility.

Let's start by colouring the t-SNE plot with sample of origin.

```{r}
tsne_plot_data <- cbind(
	reducedDim(sce, "TSNE"),
	colData(sce)
)
ggplot(tsne_plot_data, aes(x = TSNE1, y = TSNE2, colour = sample)) +
	geom_point(size = 0.2) +
	guides(
		colour = guide_legend(override.aes = list(size = 3))
	) +
	theme_classic()
```

We can also colour by cell type assignment, provided in cell metadata column 'celltype.mapped'.

This plot will requires a bit more fine tuning, due to the large number of categories:

- Use `guides()` and `guide_legend()` to control the position of the legend.
  Placing the legend under the plot works better in this case.
- Use `theme()` and its arguments `legend.key.size=` and `legend.text=` to reducing spacing and font size to fit all labels.
- Inside the quarto notebook, use `fig-height` and `fig-width` to control the overall size of the plot.

```{r}
#| fig-height: 6
#| fig-width: 8
tsne_plot_data <- cbind(
	reducedDim(sce, "TSNE"),
	colData(sce)
)
ggplot(tsne_plot_data, aes(x = TSNE1, y = TSNE2, colour = celltype.mapped)) +
	geom_point(size = 0.2) +
	guides(
		colour = guide_legend(
			position = "bottom",
			override.aes = list(size = 1)
		)
	) +
	theme_classic() +
	theme(
		legend.key.size = unit(0.3, "line"),
		legend.text = element_text(size = 7)
	)
```

### Correct batch effects

The last two plots highlight clear batch effects between subsets of samples.

Whether to remove or retain such effects - particularly sample-specific variation - depends on the intended downstream analyses. For example, when the goal is to identify equivalent cell types across samples, correcting for sample effects is typically appropriate. In contrast, analyses of developmental time courses may rely on preserving differences between otherwise similar cell types across stages of organismal development.

Correct the sample effect using `correctExperiment()`, with the following constraints:

- Set a random seed before calling `correctExperiment()` to ensure reproducibility, as some steps in the procedure are stochastic.
- Use the `batch=` argument to indicate the sample of origin for each cell.
- Provide the set of highly variable genes selected earlier to the `subset.rows=` argument.
- Configure the correction parameters using the `FastMnnParam()` function, ensuring that the three wild-type samples are merged together and the three knockout samples are merged together before the two sets are merged together.

Assign the output to a new object called `merged`.

```{r}
set.seed(10102)

merged <- correctExperiments(
    sce, 
    batch = sce$sample, 
    subset.row = chosen.hvgs,
    PARAM = FastMnnParam(
        merge.order = list(
            list(1,3,5), # WT (3 replicates)
            list(2,4,6)  # td-Tomato (3 replicates)
        )
    )
)
merged
```

If you preview the new object, you should see an assay named 'reconstructed', along a "corrected" PCA, stored alongside existing dimensionality reduction results under the name 'corrected'.

### Apply t-SNE to the corrected data

As done earlier with the uncorrected PCA, apply t-SNE to the newly corrected PCA result stored under the name 'corrected'.

```{r}
merged <- runTSNE(merged, dimred = "corrected")
merged
```

Note that this will overwrite the existing t-SNE results that was previously stored under the name 'TSNE'.

### Visually the t-SNE layout after correction

Re-use and adapt the earlier code for colouring the corrected data that you will find as a new dimensionality reduction entry under the name 'corrected'.

```{r}
tsne_plot_data <- cbind(
	reducedDim(merged, "TSNE"),
	colData(merged)
)
ggplot(tsne_plot_data, aes(x = TSNE1, y = TSNE2, colour = batch)) +
	geom_point(size = 0.2) +
	guides(
		colour = guide_legend(override.aes = list(size = 3))
	) +
	theme_classic()
```

Similarly, coloured by cell type:

```{r}
#| fig-height: 6
#| fig-width: 8
tsne_plot_data <- cbind(
	reducedDim(merged, "TSNE"),
	colData(merged)
)
ggplot(tsne_plot_data, aes(x = TSNE1, y = TSNE2, colour = celltype.mapped)) +
	geom_point(size = 0.2) +
	guides(
		colour = guide_legend(
			position = "bottom",
			override.aes = list(size = 1)
		)
	) +
	theme_classic() +
	theme(
		legend.key.size = unit(0.3, "line"),
		legend.text = element_text(size = 7)
	)
```

## Prepare for differential expression

### Aggregate cells into pseudobulks

Aggregate cells into pseudobulks using `aggregateAcrossCells()` with the following settings:

- Give the columns 'celltype.mapped' and 'sample' to identify the groups of cells that correspond to each pseudobulk

Assign the output to a new object called `summed`.

```{r}
summed <- aggregateAcrossCells(
    merged, 
    ids = colData(merged)[, c("celltype.mapped", "sample")]
)

summed
```

If you preview the new object, you should see it contains only 192 samples.
Those correspond to each of the various cell types observed in the original six samples.

### Prepare the data for edgeR

Before we worry about automating the process over every cell type, let's walk through the process for a single cell type, e.g. 'Mesenchyme'.

Create a new object called 'current' that contains only cells from the cell type of interest.

```{r}
current <- summed[, summed$celltype.mapped == "Mesenchyme"]
current
```

Next, create a `DGEList` object, which is the standard data structure for the `edgeR` package.

```{r}
y <- DGEList(counts(current), samples = colData(current))
y
```

### Check for samples with insufficient data

Look back at the object named `current`.

You will find a column of sample metadata called 'ncells'.
Check whether any sample has stricter fewer than 10 cells in its pseudobulk.
If so, remove any such sample as it would likely skew the analysis (normalisation and/or differential expression).

```{r}
discarded <- current$ncells < 10

y <- y[, !discarded]

table(discarded)
```

### Check for genes with insufficient data

We also typically want to avoid testing genes that are too lowly expressed to be reliably tested for differential expression.

To do so, use `filterByExpr()`, with the following best practices:

- Give the 'tomato' column of sample metadata to the `group=` argument.
  This forces the function to consider the number of samples in the *smallest* group when assessing whether each gene is "sufficiently" expressed to be retained.
  In some cases, a gene highly expressed in a certain condition might be excluded if that condition is only represented by a very small number of samples in the dataset.
  

```{r}
keep <- filterByExpr(y, group = current$tomato)

y <- y[keep,]

table(keep)
```

### Normalise

Having checked for unsuitable samples and genes, we can proceed to normalising the data with `calcNormFactors()`.

```{r}
y <- calcNormFactors(y)

y$samples
```

If you preview the `samples` components of the object, you should see that the column named 'norm.factors' changed from 1 everywhere to a sample-specific normalisation factor.

### Visualise the effect of normalisation

Visualise the 'Mean-Difference' plot for each sample against the others using `plotMD()`.

```{r}
#| fig-width: 8
#| fig-height: 6
par(mfrow = c(2, 3))

for (i in seq_len(ncol(y))) {
    plotMD(y, column = i)
}
par(mfrow = c(1, 1))
```

Overall, the plots above seem acceptable: centered on y=0 with a larger vertical spread at the lowest values of x.

Another way of checking whether normalisation produced sensible data is to produce an MDS plot using `limma::plotMDS()`.

Multidimensional scaling (also called principal *coordinate* analysis (PCoA)) is a dimensionality reduction technique that’s conceptually similar to principal *component* analysis (PCA).

```{r}
limma::plotMDS(
	cpm(y, log = TRUE),
	col = ifelse(y$samples$tomato, "red", "blue")
)
```

While the first dimension captures some difference between Samples 1 and 2 and the rest, the second dimension clearly separates the two conditions.

### Prepare the design matrix

The design matrix describes the experimental factors to model.

In this case, 'tomato' is the factor of interest, for which we want to compute differential expresion.
Meanwhile, 'pool' is an additional covariate that may capture - and correct for - systematic differences between pools.

Use `model.matrix()` to create a design matrix from the `y$samples` data-frame.

```{r}
design <- model.matrix(
	~ factor(pool) + factor(tomato),
	data = y$samples
)
design
```

### Estimate the overdispersion parameter

Estimate the Negative Binomial (NB) overdispersion parameter using `estimateDisp()`.

```{r}
y <- estimateDisp(y, design)

summary(y$trended.dispersion)
```

### Visualise the biological coefficient of variation

Visualise the relationship between the Biological Coefficient of Variation (BCV) and the average expression (log counts per million) for each gene, along with common and trend values, using `plotBCV()`.

```{r}
#| fig-height: 6
#| fig-width: 7
plotBCV(y)
```

### Fit a model for each gene

Fit a Quasi-Likelihood (QL) negative binomial generalized linear model for each gene using `glmQLFit()`.

The `robust = TRUE` parameter avoids distortions from outlier genes and samples.
It down-weights outliers and produces more stable and conservative dispersion estimates.

```{r}
fit <- glmQLFit(y, design, robust = TRUE)

names(fit)
```

### Visualise genewise quasi-dispersion

Plot the genewise quasi-dispersion against the gene abundance (in log2 counts per million) using `plotQLDisp()`.

```{r}
#| fig-height: 6
#| fig-width: 7
plotQLDisp(fit)
```

## Differential expression

### Perform the test

Conduct genewise quasi F-tests for the contrast of interest using `glmQLFTest()`.

```{r}
res <- glmQLFTest(fit, coef = ncol(design))

summary(decideTests(res))
```

### Access the results

Extracts the most differentially expressed genes from the test object.

**Bonus point:**
Which parameter allows you to extract more or fewer top genes?

```{r}
topTags(res)
```

## Repeat the DE test for all cell types

Thankfully, all the previous steps can be conveniently performed for each cell type in a single command, using the `pseudoBulkDGE()` function from the `scran` package.

```{r}
summed.filt <- summed[,summed$ncells >= 10]

de.results <- pseudoBulkDGE(
    summed.filt, 
    label = summed.filt$celltype.mapped,
    design = ~factor(pool) + tomato,
    coef = "tomatoTRUE",
    condition = summed.filt$tomato 
)
```

We can check that we obtain the same results as above (or try other cell types).

```{r}
cur.results <- de.results[["Mesenchyme"]]

cur.results[order(cur.results$PValue),]
```
