---
title: 'Introduction to Bioconductor'
subtitle: '<i class="fab fa-r-project"></i> for genomics data science'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2021-10-13 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='setup_and_goals.Rmd'}
```

---

# Prerequisites

- An account on the CCB cluster.

- A clone of the shared GitHub repository for this course.

## Lessons

- Introduction to `r BiocStyle::CRANpkg("ggplot2")`.

---

```{r, child='setup_and_philosophy.Rmd'}
```

---

# Exercise

## BiocManager

On the CCB cluster, all the packages required for the course are already installed.
As a result, you do not need (and should not) install any package yourself.

> Instead, you should ask the system administrators to install or update packages for you.

To give you an idea, what the system administrators did essentially comes down to:

```{r, include=TRUE, echo=TRUE, eval=FALSE}
install.packages("BiocManager")
BiocManager::install("Biostrings")
```

The package `r BiocStyle::CRANpkg("BiocManager")` is the only official Bioconductor package on CRAN.
It provides functions to install, update, and more generally manage your library of Bioconductor (and CRAN) packages.

The function `BiocManager::install()` can install packages from Bioconductor, CRAN, and even GitHub.
Here, we demonstrate it with a single package name.
However, it was used to install a whole set of Bioconductor packages and their dependencies.

- Run the function `BiocManager::valid()`.
  What does it do?

---

```{r, child='biostrings.Rmd'}
```

---

# Exercise

## Use Bioconductor packages and help pages

- Load the `r BiocStyle::Biocpkg("Biostrings")` package.

- The `r BiocStyle::Biocpkg("Biostrings")` package provides the object `DNA_ALPHABET`.
  Print the object in the console and use the help page to explain the contents of the object.

- The `Biostrings` package provides the function `luminaQuality()`.
  Run some code from the "Examples" section of the help page, and describe what the function does.

- Access the package vignette(s).
  How many vignette does the `r BiocStyle::Biocpkg("Biostrings")` package include?

- Import sequences in the file `TruSeq3-PE-2.fa`.
  What is the class of the object?

---

```{r, child='etc.Rmd'}
```