---
title: "Solution: Single-cell integration"
author: "Kevin Rue-Albrecht & Liezel Tamon"
date: "2024-04-26"
---

```{r}
#| label: load-packages
#| include: false
library(Seurat)
```

## Preamble

Refer to the vignette [Integrative analysis in Seurat v5](https://satijalab.org/seurat/articles/seurat5_integration#perform-streamlined-one-line-integrative-analysis) for guidance throughout the workflow.
Some of the functions required for this workflow are only mentioned in this vignette and not documented in the Seurat package itself.

## Import data sets to integrate

- Use `readRDS()` to import two prefiltered PBMC data sets from the files `pbmcv1_filtered.rds` and `pbmcv2_filtered.rds`.

```{r}
pbmv1 <- readRDS("data/pbmcv1_filtered.rds")
pbmv2 <- readRDS("data/pbmcv2_filtered.rds")
```

Those RDS files contain objects that have been preprocessed through quality control and filtering, as demonstrated in the Seurat day 1 lesson.

## Prepare the data sets for integration

- Use `merge()` to combine the two Seurat objects into a single object.

```{r}
obj <- merge(pbmv1, pbmv2)
```

In previous versions of Seurat, the integration workflow would have required the data to be represented as distinct Seurat objects.
When using Seurat version 5 assays, we can instead keep all the data in one object, but simply split the layers.

- Display the merged Seurat object.

```{r}
obj
```

**How do you interpet the 'layer' information?**

## Integration without correction

