---
title: "Example code for dimensionality reduction and clustering"
author: "Kevin Rue-Albrecht"
date: "01/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(cowplot)
library(reshape2)
library(Rtsne)
```

# Exercise

## Dimensionality reduction

- Download the shared folder `r/dimensionality_reduction_and_clustering/data` from the CGAT cluster.

- Import the matrix data.
  Inspect the matrix, to get familiar with rows and columns.

```{r}
mat <- read.csv("data/logcounts.csv", row.names = 1)
dim(mat)
```

- Perform PCA.
  Inspect the structure of the output.
  
```{r}
prcomp_out <- prcomp(x = t(mat), scale. = TRUE)
str(prcomp_out)
```

- Draw a scatter plot of the top two principal components and color by various experimental metadata.
  Which experimental factors show the largest variance in the data?

```{r}
cell_data <- read.csv("data/cell_metadata.csv", row.names = 1)
head(cell_data)
```

```{r}
library(ggplot2)
library(cowplot)
plot_data <- bind_cols(
  sample = rownames(prcomp_out$x),
  cell_data,
  prcomp_out$x %>% as_tibble() %>% dplyr::select(PC1, PC2)
)

gg_pca_time <- ggplot(plot_data) +
  geom_point(aes(PC1, PC2, color = Time)) +
  theme_cowplot() +
  theme(legend.position = "bottom", legend.direction = "vertical")
gg_pca_infection <- ggplot(plot_data) +
  geom_point(aes(PC1, PC2, color = Infection)) +
  theme_cowplot() +
  theme(legend.position = "bottom", legend.direction = "vertical")
gg_pca_status <- ggplot(plot_data) +
  geom_point(aes(PC1, PC2, color = Status)) +
  theme_cowplot() +
  theme(legend.position = "bottom", legend.direction = "vertical")

plot_grid(gg_pca_time, gg_pca_infection, gg_pca_status, nrow = 1)
```

- Draw a density plot, faceted by time and infection, colored by status, to inspect the variance of each experimental 
factor on the top principal component.

```{r}
plot_data <- bind_cols(
  sample = rownames(prcomp_out$x),
  cell_data,
  prcomp_out$x %>% as_tibble() %>% dplyr::select(PC1)
)

ggplot(plot_data) +
  geom_density(aes(PC1, fill = Status), color = "black", alpha = 0.5) +
  facet_grid(Time~Infection) +
  theme_cowplot()
```

- How many principal components would you keep for downstream analyses?

```{r}
scree_table <- tibble(
  PC = seq_along(prcomp_out$sdev),
  sdev = prcomp_out$sdev,
  # PC = 1:length(prcomp_out$sdev),
  percent_var = sdev^2 / sum(sdev^2),
  var_cumsum = cumsum(percent_var)
)
scree_table
```
```{r}
scree_table %>% 
  head(50) %>% 
  ggplot(aes(PC, percent_var)) +
  geom_col(fill = "grey") +
  geom_point() +
  scale_x_continuous(breaks = seq_along(scree_table$PC)) +
  labs(y = "% Variance explained", title = "Percentage of Variance Explained") +
  theme_cowplot()
```

```{r}
ggplot(head(scree_table, 200), aes(PC, var_cumsum)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 0.9) +
  scale_x_continuous(breaks = seq_along(scree_table$PC)) +
  labs(y = "Cumulative % Variance explained", title = "Cumulative Percentage of Variance Explained") +
  theme_cowplot() +
  theme(axis.text.x = element_text(size=rel(0.5)))
```

- Find the top genes associated with the top principal components.

```{r}
library(reshape2)
prcomp_out$rotation %>%
  melt(varnames = c("gene", "PC"), value.name = "loading") %>% 
  as_tibble() %>% 
  filter(PC == "PC1") %>% 
  mutate(loading_abs = abs(loading)) %>% 
  top_n(n = 20, wt = loading_abs) %>% 
  arrange(desc(loading_abs))
```

- Visualise the gene expression value of the gene with the highest loading for PC1 as color on a scatter plot of PC1 and PC2.
  What do you observe?

```{r}
plot_data <- bind_cols(
  gene = as.numeric(mat["ENSG00000172183", ]),
  cell_data,
  prcomp_out$x %>% as_tibble() %>% dplyr::select(PC1, PC2)
)

ggplot(plot_data) +
  geom_point(aes(PC1, PC2, color = gene)) +
  theme_cowplot()
```

- Compare PCA to tSNE to UMAP.
  Only use the subset of principal components selected above, when you run t-SNE and UMAP.

```{r}
library(Rtsne)
tsne_out <- Rtsne(prcomp_out$x[, 1:25])
str(tsne_out)
```
```{r}
tsne_data <- tsne_out$Y
colnames(tsne_data) <- paste0("TSNE", 1:2)

tsne_plot_data <- bind_cols(
  sample = rownames(prcomp_out$x),
  cell_data[rownames(prcomp_out$x), ] %>% as_tibble(),
  tsne_out$Y %>% as_tibble()
)

gg_tsne <- ggplot(tsne_plot_data, aes(V1, V2, color = Time)) +
  geom_point() +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  theme_cowplot() +
  theme(legend.position = "bottom", legend.direction = "vertical")

plot_grid(gg_pca_time, gg_tsne)
```

```{r}
library(umap)
umap_out <- umap(d = prcomp_out$x[, 1:25])
umap_out
```

```{r}
umap_plot_data <- bind_cols(
  sample = rownames(prcomp_out$x),
  cell_data[rownames(prcomp_out$x), ] %>% as_tibble(),
  umap_out$layout %>% as_tibble()
)

gg_umap <- ggplot(umap_plot_data, aes(V1, V2, color = Time)) +
  geom_point() +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme_cowplot() +
  theme(legend.position = "bottom", legend.direction = "vertical")

plot_grid(
  gg_pca_time,
  gg_tsne,
  gg_umap, nrow = 1)
```

