# Dimensionality reduction and clustering

## Visually extracting information from data

:::: {.columns}

::: {.column width="50%"}
### Data

<br/>

::: {.small-table}
```{r}
knitr::kable(head(assay(airway, "counts"), 20), format = "html", escape = FALSE)
```
:::
:::

::: {.column width="50%"}
### Information

```{r}
#| fig-align: center
#| fig-height: 8
#| fig-width: 7
top_var_idx <- head(order(rowVars(assay(airway, "logcounts")), decreasing = TRUE), 40)
Heatmap(
  matrix = assay(airway, "logcounts")[top_var_idx, ],
  row_names_gp = gpar(fontsize = 8),
  name = "logcounts"
)
```
:::

::::

## Sources of variation in data

Variation in data (e.g., expression) can come from multiple sources.

:::: {.columns}

::: {.column width="50%"}
### Biological

- Stimulus (e.g., infection, drug)
- Condition (e.g., healthy, disease)
- ...
:::

::: {.column width="50%"}
### Technical

- Instrumentation
- Operator
- ...
:::

::::

Whether each source of variation is seen as signal or noise depends on the goal of the study.

## Confounding variables

Experimental design is crucial to ensure that sources of interesting variation are not confounded with independent sources of uninteresting variation.

:::: {.columns}

::: {.column width="50%"}
### Confounded

<br/>

```{r}
nrow <- 8
data.frame(
  Cell = seq_len(nrow),
  Site = rep(c("S1", "S2"), each = 4),
  Treatment = rep(c("A", "B"), each = 4)
) %>%
  mutate(
    Site = cell_spec(Site, "html", color = ifelse(Site == "S2", "red", "blue")),
    Treatment = cell_spec(Treatment, "html", color = ifelse(Treatment == "A", "cyan", "orange"))
  ) %>% 
  knitr::kable(format = "html", escape = FALSE)
```
:::

::: {.column width="50%"}
### Balanced

<br/>

```{r}
nrow <- 8
data.frame(
  Cell = seq_len(nrow),
  Site = rep(c("S1", "S2"), each = 4),
  Treatment = rep(c("A", "B"), times = 4)
) %>%
  mutate(
    Site = cell_spec(Site, "html", color = ifelse(Site == "S2", "red", "blue")),
    Treatment = cell_spec(Treatment, "html", color = ifelse(Treatment == "A", "cyan", "orange"))
  ) %>% 
  knitr::kable(format = "html", escape = F)
```

:::

::::

## Feature selection

Many genes are not interesting because they don't vary much, or they
don’t have enough counts.

Filtering for feature selection is needed to:

- Select genes that display useful variation.
- Reduce memory usage and computational cost/time.

```{r}
#| fig-align: center
#| fig-height: 4
#| fig-width: 6
set.seed(1)
mat <- matrix(c(rnorm(12*4, 0, 10), rnorm(12*4, 0, 1)), nrow = 8, ncol = 12, byrow = TRUE)
ComplexHeatmap::Heatmap(mat, name = "values")
```
