# Dimensionality reduction and clustering

## Visually extracting information from data

:::: {.columns}

::: {.column width="50%"}
### Data

<br/>

::: {.small-table}
```{r}
knitr::kable(head(assay(airway, "counts"), 20), format = "html", escape = FALSE)
```
:::
:::

::: {.column width="50%"}
### Information

```{r}
#| fig-align: center
#| fig-height: 8
#| fig-width: 7
top_var_idx <- head(order(rowVars(assay(airway, "logcounts")), decreasing = TRUE), 40)
Heatmap(
  matrix = assay(airway, "logcounts")[top_var_idx, ],
  row_names_gp = gpar(fontsize = 8),
  name = "logcounts"
)
```
:::

::::

## Sources of variation in data

Variation in data (e.g., expression) can come from multiple sources.

:::: {.columns}

::: {.column width="50%"}
### Biological

- Stimulus (e.g., infection, drug)
- Condition (e.g., healthy, disease)
- ...
:::

::: {.column width="50%"}
### Technical

- Instrumentation
- Operator
- ...
:::

::::

Whether each source of variation is seen as signal or noise depends on the goal of the study.
