---
title: "Solution: Statistics, dimensionality reduction, and clustering"
author: "Author name goes here"
date: "2023-08-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Demo

### The standard normal distribution

#### pnorm()

The quantiles represent a set of possible values in the distribution.

```{r}
data_quantiles <- seq(-5, 5, 0.1)
head(data_quantiles)
```

The function `pnorm()` returns the probability of a observing a value less or equal to each quantile given.

```{r}
data_pnorm <- pnorm(q = data_quantiles) 
head(data_pnorm)
```

The base R `plot()` function can be used to quickly plot each quantile and its probability. 

```{r}
plot(data_quantiles, data_pnorm)
```

#### qnorm()

The function `qnorm()` takes as input a vector of probabilities between 0 and 1.

```{r}
data_probabilities <- seq(from = 0, to = 1, by = 0.02)
head(data_probabilities)
```

The function `qnorm()` the quantiles (i.e., values) that correspond to those probabilities.

```{r}
data_qnorm <- qnorm(p = data_probabilities) 
head(data_qnorm)
```

The base R `plot()` function can be used to quickly plot each quantile and its probability. 

```{r}
plot(data_probabilities, data_qnorm)
```

#### dnorm

The quantiles represent a set of possible values in the distribution.

```{r}
data_quantiles <- seq(-5, 5, 0.2)
head(data_quantiles)
```

The function `pnorm()` returns the probability of a observing a value less or equal to each quantile given.

```{r}
data_dnorm <- dnorm(x = data_quantiles) 
head(data_dnorm)
```

The base R `plot()` function can be used to quickly plot each quantile and its probability. 

```{r}
plot(data_quantiles, data_dnorm)
```

#### rnorm

```{r}
data_rnorm <- rnorm(n = 100, mean = 0, sd = 1)
head(data_rnorm)
```
