## Workspace management in <i class="fab fa-r-project"></i> sessions

The *workspace* is the current <i class="fab fa-r-project"></i> environment that contains all the user-defined objects (including functions) in the current session. 

As you define new objects and functions, these new objects are stored in the workspace.

When you close RStudio, you may see a message asking you "Save workspace image to .RData?"

- Agreeing saves all the objects currently in your workspace to a file on your computer;
  this is useful if you expect to restore the workspace in a later session.
- Declining discards your workspace; this is useful if you like to start <i class="fab fa-r-project"></i> with an empty, clean workspace.

<br/>

Note:

- Saving a workspace does *not* save the commands that you used to generate it.
- It is much more important to save scripts than to save workspaces.

## The working directory

The *working directory* is the location on the computer running the <i class="fab fa-r-project"></i> session relative to which commands are evaluated.

- Some commands are not affected by the working directory.

Any *relative* file path given to a command will be relative to the working directory, e.g.

```{r}
#| eval: false
#| echo: true
my_data <- read.csv(file = "data.csv")
```

Any *absolute* file path given to a command will *not* be impacted by the working directory, e.g.

```{r}
#| eval: false
#| echo: true
write.csv(x = my_data, file = "/project/obds/shared/data.csv")
```

The current working directory can be displayed using the function `getwd()`, e.g.

```{r}
#| eval: false
#| echo: true
getwd()
```

The working directory can be set using the function `setwd()`, e.g.

```{r}
#| eval: false
#| echo: true
setwd("/project/obds/shared")
```

<br/>

Note:

- The *working directory* is a location in the computer's *filesystem* (i.e., hard drive).
- The *workspace* is an environment in the computer's *memory* (i.e., RAM).

## Saving workspaces to .RData files

It is possible to store one or more objects from the workspace to an `.RData` file.

The `save()` function writes an external representation of <i class="fab fa-r-project"></i> objects to the specified file, e.g.

```{r}
#| eval: false
#| echo: true
a <- 1
b <- "A"
save(a, b, file = "workspace.RData")
```

The file can only be read using the `load()` function to restore those objects into an <i class="fab fa-r-project"></i> session.
It is not possible to read that type of file with any other program.

```{r}
#| eval: false
#| echo: true
load("workspace.RData")
```

The function `save.image()` takes only a file path and writes *all* the objects in the workspace to the specified file, e.g.

```{r}
#| eval: false
#| echo: true
a <- 1
b <- "A"
save.image(file = "workspace.RData")
```
