---
title: "Solution: Single-cell analysis using Seurat, part 2"
author: "Kevin Rue-Albrecht"
date: "2023-06-08"
date-format: long
format: html
---

```{r}
#| label: load-packages
#| include: false

library(Seurat)
```

## Exercise

### Import data

- Compare clustering results following basic normalisation and sctransform normalisation.

**Hint:** Read the file `seurat_after_qc.rds` in the shared resources.

```{r}
seurat_after_qc <- readRDS(file= "data/seurat_after_qc.rds")
```

### Default assay

Check the name of the default assay (best practice before using Seurat functions).

```{r}
DefaultAssay(seurat_after_qc)
```

### SCTransfrom

Apply SCTransfrom.

**Hint:**

- Refer to the SCTransform vignette available at <https://satijalab.org/seurat/articles/sctransform_vignette.html>.
- Consider confounding variables to specify in the argument `vars.to.regress=`.
- Consider the number of variable features to specify in the argument `variable.features.n=`.

```{r}
seurat_after_qc <- SCTransform(
  object=seurat_after_qc,
  assay="RNA",
  vars.to.regress = c("percent_mt"))
```

**Note:** The `SCTransform()` function effectively replaces the workflow performed by the functions `NormalizeData()`, `ScaleData` and `FindVariableFeatures()`.

**Where is the new normalisation stored?**

> Answer:
> Transformed data is stored in a new assay named 'SCT'.

**Note:** The 'SCT' assay is automatically set as the default assay immediately after SCTransform is run.

### Explore SCTransform results

Explore the `seurat_after_qc` object metadata and assays (best practice to get to know the object).

```{r}

```

**Can you see any change in the assays?**

> Answer:
> Yes, there is a new assay called 'SCT'.

**Can you see any new column(s) in the metadata?**

> Answer:
> There are two new columns called 'nCount_SCT. and 'nFeature_SCT'.

### Visualise SCTransform results

Use the function `VlnPlot()` to compare the total counts in the RNA and SCT assays, and to compare the number of features detected in each assay.

**Hint:** Force the function to apply the same Y-axis range to all panels in each plot, to make comparisons easier.

```{r}
# counts
VlnPlot(object=seurat_after_qc, features = c("nCount_RNA" ,"nCount_SCT") , same.y.lims = T,)
```

```{r}
# features
VlnPlot(object=seurat_after_qc, features = c("nFeature_RNA","nFeature_SCT") , same.y.lims = T)
```

**Note:**

- The original library size and number of features detected per cell are already present in the Seurat object.
- `SCTransform` adds two columns of cell metadata for library and number of features detected after SCT normalisation.

## Bonus

### Visualise a feature in the SCT assay

For the feature of your choice (e.g., `LYZ`), using the SCT assay:

- Visualise the distribution of count data.
- Visualise the distribution of log-counts in the data slot.
- Visualise the distribution of scaled log-counts.

**Hint:** For each plot, pass the output of `FetchData()` to `ggplot()`, store each plot, and combine them into a single plot using `cowplot::plot_grid()`.

```{r}
DefaultAssay(seurat_after_qc) <- "SCT"
ggplot_lyz_raw <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", slot = "counts"), aes(LYZ)) +
    geom_histogram(fill = "grey", color = "black", bins = 100) +
    coord_cartesian(ylim = c(0, 500)) +
    cowplot::theme_cowplot()
ggplot_lyz_normalised <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", slot = "data"), aes(LYZ)) +
    geom_histogram(fill = "grey", color = "black", bins = 100) +
    coord_cartesian(ylim = c(0, 500)) +
    cowplot::theme_cowplot()
ggplot_lyz_scaled <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", slot = "scale.data"), aes(LYZ)) +
    geom_histogram(fill = "grey", color = "black", bins = 100) +
    coord_cartesian(ylim = c(0, 500)) +
    cowplot::theme_cowplot()
cowplot::plot_grid(ggplot_lyz_raw, ggplot_lyz_normalised,ggplot_lyz_scaled, ncol = 1)
```

## Exercise

### Variable features

Use the function `VariableFeatures()` to display the first 10 variable genes in the SCT assay and compare to first 10 genes in the RNA assay.

**Hint:**
Look at the help page of the function to find the argument that specified which assay to access.

```{r}
VariableFeatures(seurat_after_qc, assay = "SCT")[1:10]
VariableFeatures(seurat_after_qc, assay = "RNA")[1:10]
```

**How many genes overlap?**

```{r}
length(intersect(
  VariableFeatures(seurat_after_qc, assay = "SCT")[1:10],
  VariableFeatures(seurat_after_qc, assay = "RNA")[1:10]
))
```

> Answer
> Seven genes overlap, which is more than half, so encouraging!

### Dimensionality reduction on SCT transformed data

Run a principal component analysis and UMAP on the Seurat object.

**Best practice (Seurat):** Check the Default assay (again!).

```{r}
DefaultAssay(seurat_after_qc)
```

**Hint:**

- Consider setting the argument `reduction.name=` to something other than 'pca', to preserve the PCA results based on the 'RNA' assay.
- It doesn't hurts to set the default again before important steps, even if the default assay already is the desired one, as you might sometimes run code chunks out of order.
  
```{r}
DefaultAssay(seurat_after_qc) <- "SCT"
seurat_after_qc <- RunPCA(seurat_after_qc, reduction.name = "sct.pca" )
```

Check the name of the dimensionality reduction results present in the Seurat object now (best practice to get to know the object).

```{r}
Reductions(seurat_after_qc)
```